Hydra Battles and Cie

### 2021-08-24 doc/part-hydras.tex
Alectryon: 50% of chapter6

\begin{Coqsrc}
Fixpoint gnaw (alpha : T1) (s: list nat) :=
  match s with
    | nil => alpha
    | (0::s') => gnaw  alpha s'
    | (S i :: s')  =>  gnaw (canon alpha (S i)) s'
  end.

Definition large alpha (s:list nat) := gnaw alpha s = zero.
\end{Coqsrc}

@@

\index{hydras}{Library Epsilon0!Predicates!mlarge@mlarge (minimal large sequences)}
\begin{Coqsrc}
Definition mlarge alpha (s:list nat) := path_to zero s alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Compute pp (gnaw (T1.omega * T1.omega) (interval 6 70)).
\end{Coqsrc}

\begin{Coqanswer}
 = (_omega * 2 + 56)%pT1
     : ppT1
\end{Coqanswer}

Let us try another computation.

\begin{Coqsrc}
Compute (gnaw (T1.omega * T1.omega) (interval 6 700)).
\end{Coqsrc}

\begin{Coqanswer}
 = zero : T1
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma mlarge_iff alpha x (s:list nat) :
  s <> nil -> ~ In 0 (x::s) ->
  mlarge alpha (x::s) <-> gnaw alpha (but_last x s) = one.
 \end{Coqsrc}

@@

\vspace{4pt}
 \noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets_Examples.html}{Epsilon0.Large\_Sets\_Examples}}
\begin{Coqsrc}
Example Ex1 : mlarge (T1.omega * T1.omega) (interval 6 510).
\end{Coqsrc}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets.html}{Epsilon0.Large\_Sets}}
\begin{Coqsrc}
Lemma mlarge_unicity alpha k l l' : 
  mlarge alpha (interval (S k) l) ->
  mlarge alpha (interval (S k) l') ->
  l = l'.
\end{Coqsrc}

Thus, it seems obvious that there must exist a function, parameterized by $\alpha$ which associates to any  strictly positive integer $k$ the number $l$ such that
the interval $[k,l]$ is $\alpha$-mlarge. It would be fine to write in \gallina{} a definition like this:

@@

\begin{Coqsrc}
Inductive L_spec : T1 -> (nat -> nat) -> Prop :=
| L_spec0 : forall f, (forall k, f k = k) ->  L_spec zero f
| L_spec1 : forall alpha f,
    alpha <> zero ->
    (forall k, mlarge alpha (interval (S k) (Nat.pred (f (S k))))) ->
    L_spec alpha f.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma L_spec_unicity alpha f g :
  L_spec alpha f -> L_spec alpha g -> forall k, f (S k) = g (S k).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma L_zero_inv f : L_spec zero f -> forall k, f (S k) = S k.
\end{Coqsrc}

@@

Note that our decomposition of intervals works only if the intervals we consider are not empty. In order to ensure this property, we assume that $f\;k$ is always greater than $k$, which we note \texttt{S <<= f}, or \texttt{(fun\_le S f)} (defined 
in~\href{../theories/html/hydras.Prelude.Iterates.html\#fun_le}{Prelude.Iterates}).

\begin{Coqsrc}
Definition fun_le f g  := forall n:nat, f n <=  g n.
Infix "<<=" := fun_le (at level 60).
\end{Coqsrc}

It looks also natural to show that the functions we consider are strictly monotonous. The section on successor ordinals has thus the following structure.

\begin{Coqsrc}
Section succ.
   Variables (beta : T1) (f : nat -> nat).

   Hypotheses (Hbeta : nf beta)
              (f_mono : strict_mono f)
              (f_Sle : S <<= f)
              (f_ok : L_spec beta f).

   Definition L_succ := fun k => f (S k).

   Lemma L_succ_mono : strict_mono L_succ.

   Lemma L_succ_Sle : S <<= L_succ.

   Lemma L_succ_ok : L_spec (succ beta) L_succ.

End succ.

@@

\begin{Coqsrc}
Section lim.
  Variables (lambda : T1)
            (Hnf : nf lambda)
            (Hlim : limitb lambda)
            (f : nat -> nat -> nat)
            (H : forall k, L_spec (canon lambda (S k)) (f (S k))).

  Let L_lim k := f k (S k).

  Lemma L_lim_ok : L_spec lambda L_lim.

End lim.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition L_fin i := fun k => (i + k)%nat.

Lemma L_fin_ok i : L_spec (fin i) (L_fin i).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition L_omega k := S (2 * k)%nat.

Lemma L_omega_ok : L_spec T1.omega L_omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma path_to_omega_mult (i k:nat) :
path_to (T1.omega * i) (interval (S k) (2 * (S k)))
            (T1.omega * (S i)).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma omega_mult_mlarge_0 i  : forall k,
    mlarge  (T1.omega * (S i))
            (interval (S k)
                      (Nat.pred (iterate (fun p =>  S (2 * p)%nat)
                                         (S i)
                                         (S k)))).
\end{Coqsrc}

Thus, we infer the following result:

@@

\begin{Coqsrc}
Definition L_omega_mult i (x:nat) :=  iterate L_omega i x.

Lemma L_omega_mult_ok (i: nat) :  L_spec (T1.omega * i) (L_omega_mult i).
\end{Coqsrc}

For instance, let us consider the ordinal $\omega\times 8$, and a sequence 
starting at $k=5$.

\begin{Coqsrc}
Compute L_omega_mult 8 5.
\end{Coqsrc}

\begin{Coqanswer}
= 1535
     : nat
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma L_omega_mult_eqn (i : nat) :
  forall (k : nat),  (0 < k)%nat  ->
                     L_omega_mult i k = (exp2 i * S k - 1)%nat.
\end{Coqsrc}

@@

These properties are defined in \coq{} in the library
~\href{../theories/html/hydras.Prelude.Iterates.html}{Prelude.Iterates}.

@@

\begin{Coqsrc}
(** ** Abstract properties of arithmetic functions *)

Definition strict_mono f := forall n p,  n < p -> f n < f p.

Definition strict_mono1 f := forall n p,  0 < n < p -> f n < f p.

Definition dominates_from n g f  := forall p, n <= p -> f p < g p.

Definition dominates_strong g f  := {i : nat | dominates_from i g f}.

Definition dominates g f := exists i : nat, dominates_from i g f .

Definition fun_le f g  := forall n:nat,  f n <= g n.

Infix ">>" := dominates (at level 60).

Infix ">>s" := dominates_strong (at level 60).

Infix "<<=" := fun_le (at level 60).
\end{Coqsrc}

@@

In \coq{}, we follow the  proof in~\cite{KS81}. This proof is mainly a single  proof by transfinite induction on $\alpha$ of the conjunction of the five properties.
For each $\alpha$, the three cases : $\alpha=0$, $\alpha$ is a limit, and 
$\alpha$ is a successor are considered. Inside each case, the five sub-properties are proved sequentially.

>>>>

\input{movies/snippets/Paths/gnawDef}

\noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_SetsPaths.html\#gnaw}{Epsilon0.Large\_Sets}}

\input{movies/snippets/Large_Sets/largeDef}

@@

\index{hydras}{Library Epsilon0!Predicates!mlarge@mlarge (minimal large sequences)}

\input{movies/snippets/Large_Sets/largeDef}

@@

\noindent
\emph{From Module~ \href{../theories/html/hydras.Epsilon0.Large_Sets_Examples.html\#mlarge}{Epsilon0.Large\_Sets\_Examples}}

\input{movies/snippets/Large_Sets_Examples/gnawEx1}

Let us try another computation.
\input{movies/snippets/Large_Sets_Examples/gnawEx2}

@@

\input{movies/snippets/Large_Sets/mlargeIff}

@@

\vspace{4pt}
 \noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets_Examples.html}{Epsilon0.Large\_Sets\_Examples}}

\input{movies/snippets/Large_Sets_Examples/Ex1Lemma}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets.html}{Epsilon0.Large\_Sets}}

\input{movies/snippets/Large_Sets/mlargeUnicity}

Thus, we would like to define a function, parameterized by $\alpha$ which associates to any  strictly positive integer $k$ the number $l$ such that
the interval $[k,l]$ is $\alpha$-mlarge. It would be fine to write in \gallina{} a definition like this:

@@

\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets.html}{Epsilon0.Large\_Sets}}

\input{movies/snippets/Large_Sets/LSpecDef}

@@

\input{movies/snippets/Large_Sets/LSpecUnicity}

@@

\input{movies/snippets/Large_Sets/LZeroInv}

@@

Note that our decomposition of intervals works only if the intervals we consider are not empty. In order to ensure this property, we assume that $f\;k$ is always greater than $k$, which we note \texttt{S <<= f}, or \texttt{(fun\_le S f)}.

Useful abstract properties of arithmetic functions are
defined 
in~\href{../theories/html/hydras.Prelude.Iterates.html\#fun_le}{Prelude.Iterates}.

\label{sect:abstract-arith-prop}

\input{movies/snippets/Iterates/funLeDef}

We prove also that the functions we consider are strictly monotonous. The section on successor ordinals has the following structure.

\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets.html}{Epsilon0.Large\_Sets}}

\input{movies/snippets/Large_Sets/SectionSucc}

@@

\input{movies/snippets/Large_Sets/SectionLim}

@@

\input{movies/snippets/Large_Sets/LFinDef}
\vspace{-16pt}
\input{movies/snippets/Large_Sets/LFinOk}

@@

\input{movies/snippets/Large_Sets/LOmegaDef}
\vspace{-16pt}
\input{movies/snippets/Large_Sets/LOmegaOk}

@@

\input{movies/snippets/Large_Sets/pathToOmegaMult}

@@

\input{movies/snippets/Large_Sets/omegaMultMlarge0}

@@

\input{movies/snippets/Large_Sets/LOmegaMultDef}

@@

\input{movies/snippets/Large_Sets/LOmegaMultEqn}

Correctness of the function \texttt{L\_omega\_mult} is asserted through the following lemma.

\input{movies/snippets/Large_Sets/LOmegaMultOk}

@@

@@

@@

In \coq{}, we follow the  proof in~\cite{KS81}. This proof is mainly a single  proof by transfinite induction on $\alpha$ of the conjunction of the five properties.
For each $\alpha$, the three cases : $\alpha=0$, $\alpha$ is a limit, and 
$\alpha$ is a successor are considered. Inside each case, the five sub-properties are proved sequentially, using the abstract properties defined in Sect~\vref{sect:abstract-arith-prop}
